Imputation of Incomplete Multilevel Data with R

### 2023-12-19 Manuscript/manuscript.qmd
Rename foldes

##### Clustering Inclusion
Various strategies can be employed in the imputation process to account for clustering, including the inclusion of a cluster indicator variable, conducting a separate imputation process for each cluster, or employing a simultaneous imputation method that considers clustering\cite{eddings}

The choice of strategy depends primarily on the assumptions made in the main analysis and the constraints imposed by the analyzed data. Concerning data restrictions when the analysis do not contemplate an hierarchical model (eg. descriptive way), if there are few clusters with many observations per cluster, the cluster indicator or imputation separated by groups may be appropriate \cite{graham2009}. Conversely, an imputation model that simultaneously imputes all clusters using a hierarchical model can be employed when there are more clusters or fewer observations per cluster \cite{Allison_2002}.

In a hierarchical imputation model, correlations between observations within clusters are modeled by a random effect, allowing estimation even with a limited number of observations per cluster. Multiple imputation models based on hierarchical models have been proposed, each with different assumptions. \cite{audigier}

In the particular case that the analysis contemplates the use of a hierarchical model (i.e. main model), before choosing a particular strategy, it is crucial to verify if the assumptions of the imputation model align with those of the main analysis. For example, using the cluster indicator strategy may introduce bias estimates when the model is based on a hierarchical structure \cite{taljaard2008,speidel2018}. Even if an imputation strategy congruent with the main model is preferred, it is essential to assess its appropriateness for the data, considering that less complex imputation strategies may lead to unbiased estimates in certain scenarios \cite{bailey2020}.

#### Choice of Individual Imputation Methods
The initial step involves selecting the individual imputation model for each incomplete variable in the dataset. The mice package automatically proposes imputation methods based on the variable type for 1l variables. For 2l level variables, the micemd package's () function can be utilized. This function selects among the 2l imputation methods based on the size of the clusters and the proportion of missingness in each cluster.

In addition to the package-defined imputation methods, users can specify custom methods using the "I formula". This flexibility allows for the calculation of derived variables internally or adjustments to imputation methods based on specific conditions, such as conditioning the imputation model on the level of an incomplete covariate (e.g., pregnancy test on females).

#### Model Specification
The imputation model must be congenial with the main model \cite{meng1994}. Congeniality issues arise when the imputation model and the main model make different assumptions, often due to the omission of a polynomial or interaction term or the use of transformed variables.

The imputation model can include additional terms than the main model terms that do not lead to uncongeniality. For instance, it is advisable to include the outcome variable in the imputation model for prediction variables \cite{moons2006a}. In cases where the outcome is time-to-event, the Nelson-Aalen estimate of the time to the event should be included as a covariate in the imputation model [REF]. Furthermore, the inclusion of auxiliary variables, even if not part of the main model, can be associated with the probability of missingness, enhancing the likelihood of satisfying the Missing at Random (MAR) assumption and improving estimation efficiency \cite{hardt2012a}.

The specification of imputation models is done on a variable basis, either using a prediction matrix where the type of prediction variable should be specified accordingly (see table) or through a list of formulas in the formula parameter. The latter option proves useful in formulating imputation models with polynomial terms or interactions compared to the prediction matrix specification, which requires the inclusion of additional terms as Just Another Variable (JAV) \cite{buuren2018a}.

However, even when considering an interaction term, depending on the adapted individual imputation model, it has been suggested, for instance, for treatment interaction effects, to conduct separated imputation by treatment group \cite{zhang2023}. Additionally, there are imputation models based on random forest or deep learning that can handle interaction and non-linear terms that do not require the explicit inclusion of the non-linear terms.

@@

@@

>>>>

##### Clustering Inclusion

When it comes to handling clusters during the imputation process, you've got a few options. You can use a cluster indicator variable, run separate imputation for each cluster, or go for a simultaneous imputation method that takes clustering into account \cite{eddings}.

Which strategy you choose depends on the assumptions in your main analysis and the limitations of your data. If your analysis do not use a hierarchical model (like a descriptive approach) and you have a small number of clusters with lots of observations in each, using a cluster indicator or separate imputation might be the way to go \cite{graham2009}. Conversely, if you have more clusters or fewer observations per cluster, you might want to try a simultaneous hierarchical imputation model \cite{Allison_2002}.

In a hierarchical imputation model, random effects model the correlations between observations within clusters, making it possible to estimate even with a small number of observations per cluster. There are various proposed multiple imputation models based on hierarchical models, each with its own set of assumptions \cite{audigier}.

If your analysis uses a hierarchical model, make sure the assumptions of your imputation model match up. For example, using a cluster indicator approach may lead to bias estimates if your model is based on a hierarchical structure \cite{taljaard2008,speidel2018}. Even if you prefer an imputation strategy that aligns with your main model, check if it suits your data; sometimes simpler strategies can give unbiased estimates in certain scenarios \cite{bailey2020}.

#### Choice of Individual Imputation Methods
Start by choosing the imputation model for each incomplete variable in your dataset. The mice package suggests methods based on variable types for non-clustered variables, and for the cluster ones, you can use the micemd package's find.defaultMethod() function. This function selects from different 2l imputation methods based on cluster size and the proportion of missing data in each cluster.

Besides the package-defined imputation methods, you can specify custom methods using the "I formula". This lets you calculate deterministic variables during the imputation or tweak imputation methods based on specific conditions, like conditioning the imputation model to the level of an incomplete covariate (e.g., a pregnancy test for females).

#### Model Specification

The imputation model must be congenial with the main model \cite{meng1994}. Congeniality issues arise when the imputation model and the main model make different assumptions, often due to the omission of a polynomial or interaction term or the use of transformed variables.

The imputation model can incorporate additional terms compared to the main model without causing compatibility issues. For instance, it's recommended to include the outcome variable in the imputation model for prediction variables \cite{moons2006a}. In cases where the outcome is time-to-event, the Nelson-Aalen estimate of the time to the event should be added as a covariate in the imputation model [REF]. Additionally, including auxiliary variables, even if not part of the main model, can be linked to the probability of missingness, improving the likelihood of meeting the Missing at Random (MAR) assumption and enhancing estimation efficiency \cite{hardt2012a}.

Imputation models are specified on a variable basis, either using a prediction matrix in the pred parameter or through a list of formulas in the formula parameter. In the prediction matrix option, the type of each predictor variable is specified for each incomplete variable (see table).

\begin{align}
y_{ij} =& (\beta_0 + b_{0_i})+ \beta_1x_{1_{ij}}+\beta_2x_{2_{i.}} \nonumber \\
+& (\beta_3 +  b_{3_i})x_{3_{ij}} \nonumber \\
+& \beta_4x_{4_{ij}} + \beta_{m4}\overline{x_{4_{i.}}}\nonumber \\
+& (\beta_5 +  b_{5_i})x_{5_{ij}} +\beta_{m5}\overline{x_{5_{i.}}}\nonumber \\
\end{align}

\begin{tabular}{ll}
     \toprule
Type & Definition \\
  \midrule
-2  & Cluster variable, in this case the one defined by j index  \\
 1  & Fixed variable, e.g., level-1  $x_{1_{ij}}$ or  level-2 $x_{2_{i.}}$ \\
 2  & Random variable, e.g., $x_{3_{ij}}$\\
 3  & *Fixed variable with cluster mean $x_{4_{ij}}$\\
 4  & *Random variable with cluster mean $x_{5_{ij}}$\\
-3. & Random variable only included on selection model (Heckman model)\\
-4. & Random variable only included on main model (Heckman model)\\
      \bottomrule
\end{tabular}

* 3 and 4 type have been advised to used as the inclusion of the means of the cluster is beneficial on FCS \cite{mistler2017}

Recipes have been proposed for imputing incomplete level-1 and level-2 variables for hierarchical models \cite{buuren2018a} ($\S$ 7.10), which can be convenient to follow when dealing with numerous variables (interaction terms at different levels) and models with many random effects that are prone to convergency problems due to overspecification in the imputation model. These rules were designed to ensure compatibility among the conditionally specified imputation models and congeniality between the imputation and the main model. However, they may not be applicable to all 2l imputation methods; for instance, the 2l.2stage imputation methods of the micemd package only allow the inclusion of random predictor variables (2).

On the other hand, the formula option is useful in specifying complex imputation models with polynomial terms or interactions and compared with the prediction matrix method do not requires the inclusion of additional terms as Just Another Variable (JAV) \cite{buuren2018a}($\S$ 6.4).

For some interaction terms, it has been suggested, for instance, for treatment interaction effects, to conduct separate imputation by treatment group \cite{zhang2023}. Additionally, it can be used imputation models based on random forest or deep learning that can handle interaction and non-linear terms without requiring the explicit specification of an imputation model.

@@

Given the $j$-th student belonging to the $i$-th school, the main model can be formulated as:

$$ popular_{ij} = (\beta_0 + b_i) + \beta_1sex_{ij}+ \beta_2texp_{ij}+\epsilon_{ij}$$
where $\epsilon_{ij}$ corresponds to the error term.

@@

In this dataset all the variables are level-1 (patient), except by the level-2 (study) type variable.  The analysis model for this dataset is a prediction model with `mort` as the outcome. In this tutorial we'll estimate the adjusted prognostic effect of `ct` on mortality outcomes. The estimand is the adjusted odds ratio for `ct`, after including `type`, `age` `motor_score` and `pupil`. Therefore the main model for the $i$-th patient from the $j$-th study can be described as:

$$ mort_{ij} = (\beta_0 + b_i) + \beta_1type_{.j}+ \beta_2age_{ij}+ \beta_3motorscore_{ij} + \beta_4pupil_{ij} + \beta_5ct_{ij}+\epsilon_{ij}$$
where $\epsilon_{ij}$ corresponds to the error term.
